Software implementations and analysis of three voting systems

### 2017-05-22 paper/paper.tex
Final touches on the paper

\begin{abstract}
	\noindent Various voting methods are in use across the world. They range from simple processes that everyone can understand to complex mathematical systems. This essay examines differences and similarities of the behavior of three voting methods: the Single Transferable Vote (STV), first-past-the-post (FPTP), and Schulze Method. This is done by implementing each voting method in Typescript and then running election data into them. This paper provides descriptions, pseudocode and implementations of each method. A program to generate voting results was developed in order to test the voting methods. The methods are evaluated in both singe-seat and multi-seat elections. No revolutionary results are presented, but the behavior of each method is evaluated experimentally. Several conclusions are drawn about the behavior of the methods, including to what degree the different methods provided proportionate results.
\end{abstract}
\renewcommand{\abstractname}{Referat}
\pagebreak
\begin{abstract}
  \begin{center}
  	\noindent \textbf{Mjukvaruimplementationer och analys av tre valsystem}  
  \end{center} 
  \noindent Flera valsystem används i världen. De kan vara allt ifrån enkla processer som alla kan förstå till komplexa matematiska system. Denna uppsats undersöker skillnader och likheter mellan tre valsystem: Enkel Överförbar Röst, Majoritetsval och Schulze-metoden. Detta genom att implementera alla valsystem i Typescript och testa de med olika valdata. Denna uppsats innehåller beskrivningar, pseudokod samt implementationer för alla valsystem. Ett program för att generera valdata utvecklades för att testa de olika systemen. Valsystemen testades i både enmans- och flermansval. Inga revolutionerande resultat presenteras, men varje systems beteende undersökts experimentellt. Flera slutsatser dras från de olika systemens beteende, inklusive hur proportionella resultat systemen producerade.
\end{abstract}

@@

\section{Introduction}
\subsection{Background}
Taking a collective decision as a population is difficult. To solve this issue, various voting systems with defined rules are used. They are used to show common preferences within a population, for example what politician a population wants to see elected. Several types of systems have been designed for this purpose and there are a myriad of variations of those systems \autocite{fairvoteND}. They range from simple methods such as “most votes win” to complex processes that can only be practically carried out by a computer. However, practically all voting systems are algorithmic in nature, which makes them interesting to study for both Computer Scientists and Mathematicians. When deciding on what systems that shall be used, it is useful to know what differences and similarities they have.
\subsection{Purpose}
The purpose of this essay is to evaluate differences and similarities in terms of election results in three different voting systems: Single Transferable Vote (STV), First-Past-The-Post (FPTP) and the Schulze Method. This paper will only be examining and discussing differences and similarities in the results of the methods, not touching on how practically feasible the methods might be in actual elections.
\section{Theory}
One would think that by testing all different voting systems and by setting up good criteria for how to evaluate them, it would be possible to come to a definitive conclusion of what system should be used. However, this is not the case. By setting up different mathematical criteria for how a voting system shall behave, it is impossible to satisfy all criteria simultaneously. This is what several theorems, most notably \textit{Arrow's Impossibility Theorem} show \autocite{arrow1950difficulty}. This means that any decision regarding choice of electoral system is a balance of their strengths and weaknesses. Simply put: there is no perfect voting system and never will be. This does, however, not mean that all voting systems are equally efficient, elegant or accurate.

@@

All voting algorithms have been implemented in Typescript – mostly due to the author having previous experience with JavaScript. Typescript is a superset of JavaScript with multiple extra features. Most importantly, Typescript has optional static type checking. Sample code written in Typescript can be seen in code example~\ref{lst:typescript example}. The full source code with instructions for how to run the code can be found in Appendix~\ref{sec:conde implementation}.
\codeblock{code/typescriptexample.ts}{lst:typescript example}{Basic Typescript syntax}

@@

In order to test the methods and their implementations, test data is needed. There are cases where extensive election data is published, such as in Maltese elections, but full ballot data, which is needed as input for the implementations used in this paper has not been found. Instead, the algorithms will be tested with data generated from a simple ballot-generating computer program. The program used for this purpose in this paper is highly primitive, as modeling preferences within a population is far beyond the scope of this paper. The full ballot generation program can be found in the \Code{/generator} folder. The goal of the program is to produce ballots that rank every candidate in order of preference (see figure~\ref{fig:stv ballot} on page~\pageref{fig:stv ballot}). The program incorporates the idea of an \textit{ideology} where each candidate is assigned one. Each ideology has has a list of weighted preferences indicating how a voter for this ideology is likely to vote. The weighed preferences are generated via an abstraction. A single configuration object is used to create all ideologies.
\codeblock{code/generator/ideologyConfig.ts}{lst:ideology configuration}{Configuration object for creating ideologies}

@@

The scenario is built around electing one candidate out of eight. The spread of votes within ideologies is ``flat'' and all ideologies are so called ``weak''. This means that the general spread of votes between candidates will be even. An example of the distribution of first-preference votes generated in this scenario can be seen in figure~\ref{fig:example of scenario 1}. The full configuration object can be seen in code example~\ref{lst:scenario 1}.
\codeblock{code/generator/scenario1.ts}{lst:scenario 1}{Configuration object for scenario 1}

@@

The second scenario elects 3 candidates out of 9 total. The parameters are set to produce a ``normal'' spread of votes between candidates. An example distribution of first-preference votes generated in this scenario can be seen in figure~\ref{fig:example of scenario 2}. The full configuration can be seen in code example~\ref{lst:scenario 2}.
\codeblock{code/generator/scenario2.ts}{lst:scenario 2}{Scenario 2}

@@

The second scenario elects 3 candidates out of 9 total. It features a steep spread of votes between candidates. An example distribution of first-preference votes generated in this scenario can be seen in figure~\ref{fig:example of scenario 3}. The full configuration can be seen in code example~\ref{lst:scenario 3}.
\codeblock{code/generator/scenario3.ts}{lst:scenario 3}{Scenario 3}

@@

The standard deviation of the outcomes represents the variance in outcome due to the pseudorandom nature of the generator program outlined in Section~\ref{sec:vote-generation}. The standard deviation is calculated according to equation~\ref{eq:standard-deviation}. 

@@

The first-past-the-post voting method is a method designed for electing one or several (N) candidates out of a set of candidates. Each voter has one vote that may be allocated to any candidate. In a FPTP election the N candidates with most votes get elected \autocite{wilkinson2017}.

@@

The first-past-the-post program is implemented in a single file (\Code{/fptp/index.ts}). The inputs of the program can be seen in code example~\ref{lst:fptp inputs}.
\codeblock{code/fptp/inputs.ts}{lst:fptp inputs}{Inputs for FPTP program}

@@

The Single Transferable Vote (STV) is a more complicated voting system than the first-past-the-post method \autocite{electotal-reform-society2015}. STV is a multi-seat election method, although it can be used for single-seat elections too. STV becomes equivalent to the Instant-runoff voting method (IRV) in single-seat elections. In an STV election, voters are given ballots in which they are supposed to rank candidates in order. The specific rules for wether one needs to rank all candidates or just a few can vary. For the sake of simplicity, all candidates must be ranked at unique and linear values in this implementation. An example of an STV ballot can be seen in figure~\ref{fig:stv ballot}. By having voters convey more information in a ballot, there is more data available that the method can take advantage of. 

@@

	\includegraphics[height=140px]{ballot}
	\caption{Sample STV ballot}

@@

Using this data, votes are transferred between candidates within a ballot, hence the name Single Transferable Vote. The method relies on the concept of a quota, which is the number of votes one needs to be elected. The Droop-Quota, defined as $(\frac{\text{total valid poll}}{\text{seats} + 1})+1$, is often used and is the one implemented in this paper. If a candidate receives equal or more votes than the quota, he/she is elected. If the number of votes for a candidate exceed the quota, a fraction of his/her votes are transferred to the next choice on the ballots that voted for that particular candidate. This process is repeated until there are no candidates with more votes than the quota. If there still are seats yet to be filled, the candidate with fewest votes is eliminated and his/her votes are transferred to the next choice on the ballots which supported the candidate. This process is repeated until all seats are filled \autocite{king200}. The process can become quite complex with many ballots and transferring of fractional votes. A computer is therefore essential to resolve large scale elections.

@@

The STV algorithm uses a tree structure to represent the state of the election. Every candidate is a top-level node with an attribute showing the number of current votes for the candidate. Each node has child nodes representing the next preferences of the voters for that particular candidate, see figure~\ref{fig:tree structure}. It is easy to manipulate branches via recursive functions. Merging and multiplying branches is used when transferring votes from one candidate node to another. A candidate node is a Typescript class. It can be found in the \Code{stv/candidate.ts} file and has the following properties and functions:

@@

	\includegraphics[height=200px]{Tree}
	\caption{Simplified and shortened version of the tree structure. Each candidate is a top-level node with children, signifying the preferences of the voters for that candidate.}

@@

\codeblock{code/stv/candidateNode.ts}{lst:stv candidate node}{The CandidateNode class}

@@

\codeblock{code/stv/distribute.ts}{lst:stv distribute function}{The distribute function}
In order to eliminate and transfer surplus votes from any candidate node $A$, the children of $A$ are passed into the distribute function. In the case of transferring surplus votes, the votes for each child node is first multiplied by $\frac{\text{surplus votes}}{\text{total votes}}$ as outlined in the pseudocode in Section~\ref{alg:stv psuedocode}. The full source code for functions that manipulate the tree structure can be found in \Code{/stv/tree.ts}.

@@

The Schulze Method is a method used mainly to determine results in single-winner elections but can also be used to provide rankings and elect multiple candidates in a single election \autocite{Schulze2011}. Ballots are identical to those of the Single Transferable Vote (see figure~\ref{fig:stv ballot} on page~\pageref{fig:stv ballot}). The method is compliant with the Condorcet Criterion, which means that if there is a candidate who the majority prefers in a pairwise comparison with every other candidate, that candidate  \autocite{johnson2005}. As mentioned, this method relies on comparing the preferences of every candidate to one another as shown in table~\ref{tab:pairwise comparison matrix}. If there is a Condorcet winner the process is simple: declare the Condorcet winner a winner and run another iteration without that candidate. Repeat this process until all seats are filled. However, there can be Condorcet ties, where there is no Condorcet winner. There are multiple methods used for resolving Condorcet ties, with one of them being the Schulze Method.

The Schulze Method resolves ties by investigating possible paths between candidates. For example, if out of a total of 50 ballots 30 ballots prefer $B>A$, 28 ballots prefer $A>C$ and 27 ballots prefer $C>B$, a path from $A$ to $B$ can be created by going $A \rightarrow C \rightarrow B$. The path is said to have the $strength$ of the weakest link in the path. In this example, the path strength between A and B is 27 due to the link between $C$ and $B$ having a strength of 27. There may be several paths between two candidates and the goal of the process is to find the strongest path between any candidate $A$ and $B$, written as $p[A,B]$. By finding the strongest path between all nodes a result can be obtained where $p[X,Y] \geq p[Y,X]$ which means that candidate $X$ wins. The Schulze Method also provides a linear ranking between all candidates, for example that $E > B > A > C > D$. By selecting the $N$ top candidates one can use the method for a multi-seat election. As the process can become very complex as the number of candidates grow, a computer is needed to resolve large elections.

The difficult problem in this method is finding the strongest path between candidates. The problem is known as the widest path problem in graph theory. An efficient and relatively simple way to compute this problem is via the Floyd-Warshall algorithm which is used in the implementation \autocite{wikipedia2017warshall}. See Code Example~\ref{lst:schulze algorithm} for the full algorithm.

@@

\caption{Pairwise comparison matrix used in the Schulze method. For example, 17 ballots prefer $A>B$ and 33 ballots prefer $B>A$.}

@@

 & \multicolumn{1}{l|}{A} & \multicolumn{1}{l|}{B} & \multicolumn{1}{l|}{C} & \multicolumn{1}{l|}{D} & \multicolumn{1}{l|}{E} \\ \hline
\multicolumn{1}{|l|}{A} & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}17 & \cellcolor[HTML]{FFDDDD}14 & \cellcolor[HTML]{DDFFDD}35 & \cellcolor[HTML]{DDFFDD}30 \\ \hline
\multicolumn{1}{|l|}{B} & \cellcolor[HTML]{DDFFDD}33 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}24 & \cellcolor[HTML]{DDFFDD}47 & \cellcolor[HTML]{DDFFDD}36 \\ \hline
\multicolumn{1}{|l|}{C} & \cellcolor[HTML]{DDFFDD}36 & \cellcolor[HTML]{DDFFDD}26 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{DDFFDD}40 & \cellcolor[HTML]{DDFFDD}42 \\ \hline
\multicolumn{1}{|l|}{D} & \cellcolor[HTML]{FFDDDD}15 & \cellcolor[HTML]{FFDDDD}3 & \cellcolor[HTML]{FFDDDD}10 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}18 \\ \hline
\multicolumn{1}{|l|}{E} & \cellcolor[HTML]{FFDDDD}20 & \cellcolor[HTML]{FFDDDD}14 & \cellcolor[HTML]{FFDDDD}8 & \cellcolor[HTML]{DDFFDD}32 & \cellcolor[HTML]{9B9B9B} \\ \hline

@@

The Schulze Method is implemented within the file \Code{schulze/index.ts}. It takes two inputs: a list of ballots and the number of seats to be elected. A ballot is an ordered list of candidates. The output of the program is a list of winners. Data representation can be seen in Code Example~\ref{lst:schulze-maps}.
\codeblock{code/schulze/maps.ts}{lst:schulze-maps}{Data representation}
The program then runs the algorithm outlined in the pseudocode in Section~\ref{alg:schulze psuedocode}.
\codeblock{code/schulze/algorithm.ts}{lst:schulze algorithm}{Main algorithm}
\section{Evaluation of results}

@@

Scenario 1 featured eight candidates where one was to be elected. Vote spread was flat, meaning that votes tended to be diverse in their preferences. This means that there is not an obvious winner present. This can be seen from the first-preference votes in the scenario seen in figure~\ref{fig:example of scenario 1} on Page~\pageref{fig:example of scenario 1}. The full result in this scenario can be seen in figure~\ref{fig:scenario 1 results}. The outlier in this scenario seems to be the FPTP method, electing candidate 0 more often than both other methods. Key metrics for the results in this scenario can be found in table~\ref{tab:scenario 1 result}. The misrepresentation metric show that FPTP is not entirely representative. The STV and Schulze method seem to produce very similar results.

@@

	\caption{Results in scenario 1}

@@

	FPTP preformed considerably worse at being representative than the other methods. This was true for both close single- and multi-seat elections, excluding The Schulze Method in scenario 2 (see Point~\ref{num:conclusion-schulze}). At the core, both the STV and Schulze Method have more data to use for determining winners than what FPTP does. The FPTP method only has access to first-preference votes which does not convey as much information as a ballot listing all candidates in order of preference does.
	\item STV seems to be the most proportional method out of the 3 tested, but is not without its flaws.

	STV provided the most proportionate results overall of all three methods tested. However, as it goes with voting system, it has several flaws. It is for example possible to increase the chance of a candidate winning by lowering him/her on a ballot, and vice versa. STV is the only multi-seat proportional voting system tested in this paper, and the results obtained reinforce its position as just that.
\end{enumerate}
While there is no perfect voting system, they all preform differently. From the conclusions outlined above, a decision can be taken about what system analyzed in this paper to be used. Generally, if a proportional representation is desired (which in a democracy it usually is), STV will likely be the most appropriate method to use. In single-seat elections, the Schulze Method should also be considered. FPTP is the simplest method to both use and understand, but cannot be considered a proportionate voting method.

@@

\printbibliography[
heading=bibintoc,
title={References}
]
\pagebreak
\begin{appendices}
\section{Code implementation}
\label{sec:conde implementation}
\begin{figure}[H]
\dirtree{%
.1 /.
.2 fptp/.
.3 index.ts.
.2 generator/.
.3 ballots.ts.
.3 index.ts.
.3 candidates.ts.
.3 interfaces.ts.
.2 schulze/.
.3 index.ts.
.2 stv/.
.3 candidate.ts.
.3 tree.ts.
.3 election.ts.
.3 index.ts.
.2 .gitignore.
.2 LICENSE.txt.
.2 package.json.
.2 README.md.
.2 tsconfig.json.
.2 tslint.json.
}
\caption{Project structure}
\label{fig:project-structure}
\end{figure}
The full source code for this project can be found at \url{https://github.com/adriansalamon/gymnasiearbete}. It is open sourced under the MIT License. In order to run any of the implementations, the Typescript files (\Code{.ts}) need to be compiled to JavaScript files (\Code{.js}) using the Typescript Compiler. Then the code can be executed using Node.js. Code Example~\ref{lst:project-example} shows an example of how to run an election.
\codeblock{code/projectExample.ts}{lst:project-example}{Example of how to run a Schulze Election with the code.}
All source code is commented and includes extensive use type annotations to help with orientation. The full project structure can be seen in Figure~\ref{fig:project-structure}.

>>>>

\begin{abstract}
	\noindent Various voting methods are in use across the world. They range from simple processes that everyone can understand to complex mathematical systems. This essay examines differences and similarities of the behavior of three voting methods: the single transferable vote (STV), first-past-the-post (FPTP), and Schulze method. This is done by implementing each voting method in Typescript, a programming language, and then running election data into them. This paper provides descriptions, pseudocode and implementations of each method. A program to generate voting results was developed in order to test the voting methods. The methods are evaluated in both singe-seat and multi-seat elections. No new findings are presented, but the behavior of each method is evaluated experimentally. Several conclusions are drawn about the behavior of the methods, including to what degree the different methods provided proportionate results.
\end{abstract}
\renewcommand{\abstractname}{Sammanfattning}
\pagebreak
\begin{abstract}
  \begin{center}
  	\noindent \textbf{Mjukvaruimplementationer och analys av tre valsystem}  
  \end{center} 
  \noindent Flera valsystem används i världen. De kan vara allt ifrån enkla processer som alla kan förstå till komplexa matematiska system. Denna uppsats undersöker skillnader och likheter mellan tre valsystem: Enkel Överförbar Röst, Majoritetsval och Schulze-metoden. Detta genom att implementera alla valsystem i programmeringsspråket Typescript och testa de med olika valdata. Denna uppsats innehåller beskrivningar, pseudokod samt implementationer för alla valsystem. Ett program för att generera valdata utvecklades för att testa de olika systemen. Valsystemen testades i både enmans- och flermansval. Inga nya upptäckter presenteras, men varje systems beteende undersökts experimentellt. Flera slutsatser dras från de olika systemens beteende, inklusive hur proportionella resultat systemen producerade.
\end{abstract}

@@

\section{Introduction}
\subsection{Background}
Taking a collective decision as a population is difficult. To solve this issue, various voting systems with defined rules are used. They are used to show common preferences within a population, for example what politician a population wants to see elected. Several types of systems have been designed for this purpose and there are a myriad of variations of those systems \autocite{fairvoteND}. They range from simple methods such as “most votes win” to complex processes that can only be practically carried out by a computer. However, practically all voting systems are algorithmic in nature, which makes them interesting to study for both Computer Scientists and Mathematicians. When deciding on what systems that shall be used, it is useful to know what differences and similarities they have.
\subsection{Purpose}
The purpose of this essay is to evaluate differences and similarities in terms of election results in three different voting systems: single transferable vote (STV), first-past-the-post (FPTP) and the Schulze method. This paper will only be examining and discussing differences and similarities in the results of the methods, not touching on how practically feasible the methods might be in actual elections.
\section{Theory}
One would think that by testing all different voting systems and by setting up good criteria for how to evaluate them, it would be possible to come to a definitive conclusion of what system should be used. However, this is not the case. By setting up different mathematical criteria for how a voting system shall behave, it is impossible to satisfy all criteria simultaneously. This is what several theorems, most notably \textit{Arrow's Impossibility Theorem} show \autocite{arrow1950difficulty}. This means that any decision regarding choice of electoral system is a balance of their strengths and weaknesses. Simply put: there is no perfect voting system and never will be. This does, however, not mean that all voting systems are equally efficient, elegant or accurate.

@@

All voting algorithms have been implemented in Typescript – mostly due to the author having previous experience with JavaScript. Typescript is a programming language developed my Microsoft. The language is superset of JavaScript with multiple extra features. Most importantly, Typescript has optional static type checking. Sample code written in Typescript can be seen in Code Example~\ref{lst:typescript example}. The full source code with instructions for how to run the code can be found in Appendix~\ref{sec:conde implementation}.
\codeblock{code/typescriptexample.ts}{lst:typescript example}{Basic Typescript syntax.}

@@

In order to test the methods and their implementations, test data is needed. There are cases where extensive election data is published, such as in Maltese elections, but full ballot data, which is needed as input for the implementations used in this paper has not been found. Instead, the algorithms will be tested with data generated from a simple ballot-generating computer program. The program used for this purpose in this paper is highly primitive, as modeling preferences within a population is far beyond the scope of this paper. The full ballot generation program can be found in the \Code{generator/} folder. The goal of the program is to produce ballots that rank every candidate in order of preference (see Figure~\ref{fig:stv ballot} on page~\pageref{fig:stv ballot}). The program incorporates the idea of an \textit{ideology} where each candidate is assigned one. Each ideology has has a list of weighted preferences indicating how a voter for this ideology is likely to vote. The weighed preferences are generated via an abstraction. A single configuration object is used to create all ideologies.
\codeblock{code/generator/ideologyConfig.ts}{lst:ideology configuration}{Configuration object for creating ideologies. Allows for changing the variables to produce different types of electorates and election circumstances. Used to test the methods in different circumstances.}

@@

The scenario is built around electing one candidate out of eight. The spread of votes within ideologies is ``flat'' and all ideologies are so called ``weak''. This means that the general spread of votes between candidates will be even. An example of the distribution of first-preference votes generated in this scenario can be seen in Figure~\ref{fig:example of scenario 1}. The full configuration object can be seen in Code Example~\ref{lst:scenario 1}.
\codeblock{code/generator/scenario1.ts}{lst:scenario 1}{Configuration object for scenario 1. It features 8 candidates, 1 seat to be elected, and a relatively even spread of votes.}

@@

The second scenario elects 3 candidates out of 9 total. The parameters are set to produce a ``normal'' spread of votes between candidates. An example distribution of first-preference votes generated in this scenario can be seen in Figure~\ref{fig:example of scenario 2}. The full configuration can be seen in Code Example~\ref{lst:scenario 2}.
\codeblock{code/generator/scenario2.ts}{lst:scenario 2}{Configuration object for scenario 2. It features 9 candidates, 3 seats to be elected, and a ``normal'' spread of votes.}

@@

The third scenario elects 3 candidates out of 9 total. It features a steep spread of votes between candidates. An example distribution of first-preference votes generated in this scenario can be seen in Figure~\ref{fig:example of scenario 3}. The full configuration can be seen in Code Example~\ref{lst:scenario 3}.
\codeblock{code/generator/scenario3.ts}{lst:scenario 3}{Configuration object for scenario 3. It features 9 candidates, 3 seats to be elected, and a steep spread of votes.}

@@

The standard deviation of the outcomes represents the variance in outcome due to the pseudorandom nature of the generator program outlined in Section~\ref{sec:vote-generation}. The standard deviation is calculated according to Equation~\ref{eq:standard-deviation}. 

@@

The first-past-the-post voting method is a method designed for electing one or several (N) candidates out of a set of candidates. Each voter has one vote that may be allocated to any candidate. In a FPTP election the N candidates with most votes get elected. \autocite{wilkinson2017}

@@

The first-past-the-post program is implemented in a single file (\Code{fptp/index.ts}). The inputs of the program can be seen in Code Example~\ref{lst:fptp inputs}.
\codeblock{code/fptp/inputs.ts}{lst:fptp inputs}{Inputs for FPTP program.}

@@

The single transferable vote (STV) is a more complicated voting system than the first-past-the-post method. STV is a multi-seat election method, although it can be used for single-seat elections too. STV becomes equivalent to the Instant-runoff voting method (IRV) in single-seat elections. In an STV election, voters are given ballots in which they are supposed to rank candidates in order. The specific rules for wether one needs to rank all candidates or just a few can vary. For the sake of simplicity, all candidates must be ranked at unique and linear values in this implementation. An example of an STV ballot can be seen in Figure~\ref{fig:stv ballot}. By having voters convey more information in a ballot, there is more data available that the method can take advantage of. \autocite{electotal-reform-society2015}

@@

	\input{images/ballot.tex}
	\caption{Sample STV ballot. A voter fills in all candidates in order of preference.}

@@

Using this data, votes are transferred between candidates within a ballot, hence the name single transferable vote. The method relies on the concept of a quota, which is the number of votes one needs to be elected. The Droop-Quota, defined as $(\frac{\text{total valid poll}}{\text{seats} + 1})+1$, is often used and is the one implemented in this paper. If a candidate receives equal or more votes than the quota, he/she is elected. If the number of votes for a candidate exceed the quota, a fraction of his/her votes are transferred to the next choice on the ballots that voted for that particular candidate. This process is repeated until there are no candidates with more votes than the quota. If there still are seats yet to be filled, the candidate with fewest votes is eliminated and his/her votes are transferred to the next choice on the ballots which supported the candidate. This process is repeated until all seats are filled. The process can become quite complex with many ballots and transferring of fractional votes. A computer is therefore essential to resolve large scale elections. \autocite{king200}

@@

The STV algorithm uses a tree structure to represent the state of the election. Every candidate is a top-level node with an attribute showing the number of current votes for the candidate. Each node has child nodes representing the next preferences of the voters for that particular candidate, see Figure~\ref{fig:tree structure}. It is easy to manipulate branches via recursive functions. Merging and multiplying branches is used when transferring votes from one candidate node to another. A candidate node is a Typescript class. It can be found in the \Code{stv/candidate.ts} file and has the following properties and functions:

@@

	\input{images/tree.tex}
	\caption{Simplified and shortened version of the tree structure. Each candidate is a top-level node with child-nodes, signifying the next preferences of the voters for that candidate. The tree represents the full state of the election.}

@@

\codeblock{code/stv/candidateNode.ts}{lst:stv candidate node}{The CandidateNode class. Includes a list of child nodes and methods to manipulate the node.}

@@

\codeblock{code/stv/distribute.ts}{lst:stv distribute function}{The distribute function. Main method used to manipulate the state of the election.}
In order to eliminate and transfer surplus votes from any candidate node $A$, the children of $A$ are passed into the distribute function. In the case of transferring surplus votes, the votes for each child node is first multiplied by $\frac{\text{surplus votes}}{\text{total votes}}$ as outlined in the pseudocode in Section~\ref{alg:stv psuedocode}. The full source code for functions that manipulate the tree structure can be found in \Code{stv/tree.ts}.

@@

The Schulze method is a method used mainly to determine results in single-winner elections but can also be used to provide rankings and elect multiple candidates in a single election. Ballots are identical to those of the single transferable vote (see Figure~\ref{fig:stv ballot} on page~\pageref{fig:stv ballot}). The method is compliant with the Condorcet Criterion, which means that if there is a candidate who the majority prefers in a pairwise comparison with every other candidate, that candidate will win  As mentioned, this method relies on comparing the preferences of every candidate to one another as shown in Table~\ref{tab:pairwise comparison matrix}. If there is a Condorcet winner the process is simple: declare the Condorcet winner a winner and run another iteration without that candidate. Repeat this process until all seats are filled. However, there can be Condorcet ties, where there is no Condorcet winner. There are multiple methods used for resolving Condorcet ties, with one of them being the Schulze method. \autocite{johnson2005}

The Schulze method resolves ties by investigating possible paths between candidates. For example, if out of a total of 50 ballots 30 ballots prefer $B>A$ ($A$ over $B$), 28 ballots prefer $A>C$ and 27 ballots prefer $C>B$, a path from $A$ to $B$ can be created by going $A \rightarrow C \rightarrow B$. The path is said to have the $strength$ of the weakest link in the path. In this example, the path strength between A and B is 27 due to the link between $C$ and $B$ having a strength of 27. There may be several paths between two candidates and the goal of the process is to find the strongest path between any candidate $A$ and $B$, written as $p[A,B]$. By finding the strongest path between all nodes a result can be obtained where $p[X,Y] \geq p[Y,X]$ which means that candidate $X$ wins. The Schulze method also provides a linear ranking between all candidates, for example that $E > B > A > C > D$. By selecting the $N$ top candidates one can use the method for a multi-seat election. As the process can become very complex as the number of candidates grow, a computer is needed to resolve large elections. \autocite{Schulze2011}

The difficult problem in this method is finding the strongest path between candidates. The problem is known as the widest path problem in graph theory. An efficient and relatively simple way to compute this problem is via the Floyd-Warshall algorithm which is used in the implementation \autocite{wikipedia2017warshall}. See Code Example~\ref{lst:schulze algorithm} on page~\pageref{lst:schulze algorithm} for the full algorithm.

@@

\caption{Pairwise comparison matrix used in the Schulze method. $d$[$X$,$Y$] detonates the number of ballots that prefer candiate $X$ over $Y$.}

@@

 & \multicolumn{1}{l|}{\small $d$[$*$,$A$]} & \multicolumn{1}{l|}{\small $d$[$*$,$B$]} & \multicolumn{1}{l|}{\small $d$[$*$,$C$]} & \multicolumn{1}{l|}{\small $d$[$*$,$D$]} & \multicolumn{1}{l|}{\small $d$[$*$,$E$]} \\ \hline
\multicolumn{1}{|l|}{\small $d$[$A$,$*$]} & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}17 & \cellcolor[HTML]{FFDDDD}14 & \cellcolor[HTML]{DDFFDD}35 & \cellcolor[HTML]{DDFFDD}30 \\ \hline
\multicolumn{1}{|l|}{\small $d$[$B$,$*$]} & \cellcolor[HTML]{DDFFDD}33 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}24 & \cellcolor[HTML]{DDFFDD}47 & \cellcolor[HTML]{DDFFDD}36 \\ \hline
\multicolumn{1}{|l|}{\small $d$[$C$,$*$]} & \cellcolor[HTML]{DDFFDD}36 & \cellcolor[HTML]{DDFFDD}26 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{DDFFDD}40 & \cellcolor[HTML]{DDFFDD}42 \\ \hline
\multicolumn{1}{|l|}{\small $d$[$D$,$*$]} & \cellcolor[HTML]{FFDDDD}15 & \cellcolor[HTML]{FFDDDD}3 & \cellcolor[HTML]{FFDDDD}10 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}18 \\ \hline
\multicolumn{1}{|l|}{\small $d$[$E$,$*$]} & \cellcolor[HTML]{FFDDDD}20 & \cellcolor[HTML]{FFDDDD}14 & \cellcolor[HTML]{FFDDDD}8 & \cellcolor[HTML]{DDFFDD}32 & \cellcolor[HTML]{9B9B9B} \\ \hline

@@

The Schulze method is implemented within the file \Code{schulze/index.ts}. It takes two inputs: a list of ballots and the number of seats to be elected. A ballot is an ordered list of candidates. The output of the program is a list of winners. Data representation can be seen in Code Example~\ref{lst:schulze-maps}.
\codeblock{code/schulze/maps.ts}{lst:schulze-maps}{Data representation in the Schulze algorithm.}
The program uses these inputs and runs the algorithm outlined in the pseudocode in Section~\ref{alg:schulze psuedocode}. The Typescript implementation of this algorithm is detailed in Code Example~\ref{lst:schulze algorithm}.
\codeblock{code/schulze/algorithm.ts}{lst:schulze algorithm}{Main algorithm for calculating the outcome in the Schulze method. Variant of the Floyd-Warshall algorithm.}
\section{Results}

@@

Scenario 1 featured eight candidates where one was to be elected. Vote spread was flat, meaning that votes tended to be diverse in their preferences. This means that there is not an obvious winner present. This can be seen from the first-preference votes in the scenario seen in Figure~\ref{fig:example of scenario 1} on page~\pageref{fig:example of scenario 1}. The full result in this scenario can be seen in Figure~\ref{fig:scenario 1 results}. The outlier in this scenario seems to be the FPTP method, electing candidate 0 more often than both other methods. Key metrics for the results in this scenario can be found in Table~\ref{tab:scenario 1 result}. The misrepresentation metric show that FPTP is not entirely representative. The STV and Schulze method seem to produce very similar results.

@@

	\caption{Results in scenario 1.}

@@

	FPTP preformed considerably worse at being representative than the other methods. This was true for both close single- and multi-seat elections, excluding The Schulze method in scenario 2 (see point~\ref{num:conclusion-schulze}). At the core, both the STV and Schulze method have more data to use for determining winners than what FPTP does. The FPTP method only has access to first-preference votes which does not convey as much information as a ballot listing all candidates in order of preference does.
	\item STV seems to be the most proportional method out of the 3 tested, but is not without its flaws.

	STV provided the most proportionate results overall of all three methods tested. However, as it goes with voting system, it has several flaws. It is for example possible to increase the chance of a candidate winning by lowering him/her on a ballot, and vice versa. STV is the only multi-seat proportional voting system tested in this paper, and the results obtained reinforce its position as just that.
	\item In elections where the population has clear preferences, the three methods provide similar results.

	In election scenario 3, from which results can be seen in Section~\ref{sec:resuls-scenario3}, the three methods tested produced identical results. This scenario feautred a steep spread of votes wich meant that the preferences of the population were very clear. This can be seen from the distribution of first-preference votes in this scenario, seen in Figure~\ref{fig:example of scenario 3} on page \pageref{fig:example of scenario 3}. From this, it can be inferred that the results from the different methods converge as the electorate has clearer preferences.
\end{enumerate}
While there is no perfect voting system, they all perform differently. From the conclusions outlined above, a decision can be taken about what system analyzed in this paper to be used. Generally, if a proportional representation is desired (which in a democracy it usually is), STV will likely be the most appropriate method to use. In single-seat elections, the Schulze method should also be considered. FPTP is the simplest method to both use and understand, but cannot be considered a proportionate voting method.

@@

\printbibliography[
heading=bibintoc,
title={References}
]
\pagebreak
\begin{appendices}
\section{Code implementation}
\label{sec:conde implementation}
\begin{figure}[H]
\dirtree{%
.1 /.
.2 fptp/.
.3 index.ts.
.2 generator/.
.3 ballots.ts.
.3 index.ts.
.3 candidates.ts.
.3 interfaces.ts.
.2 schulze/.
.3 index.ts.
.2 stv/.
.3 candidate.ts.
.3 tree.ts.
.3 election.ts.
.3 index.ts.
.2 .gitignore.
.2 LICENSE.txt.
.2 package.json.
.2 README.md.
.2 tsconfig.json.
.2 tslint.json.
}
\caption{Project structure.}
\label{fig:project-structure}
\end{figure}
The full source code for this project can be found at \url{https://github.com/adriansalamon/gymnasiearbete}. It is open sourced under the MIT License. In order to run any of the implementations, the Typescript files (\Code{.ts}) need to be compiled to JavaScript files (\Code{.js}) using the Typescript Transpiler. Then the code can be executed using Node.js. Code Example~\ref{lst:project-example} shows an example of how to run an election.
\codeblock{code/projectExample.ts}{lst:project-example}{Example of how to run a Schulze Election with the code.}
All source code is commented and includes extensive use type annotations to help with orientation. The full project structure can be seen in Figure~\ref{fig:project-structure}.
